---
title: "Individual EDA: Evaluating Demographic Disparities in Pulse Oximetry Accuracy"
subtitle: "OpenOximetry Discharge Data"
author: "Patricia Escobar Contreras"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
editor: 
  markdown: 
    wrap: 72
---

## Overview

This EDA covers the **OpenOximetry** dataset:

-   **OpenOximetry** (UCSF Hypoxia Lab) - Used to audit the clinical
    reliability of pulse oximeters by comparing their digital readings
    against the "gold standard" of direct blood oxygen measurements.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(visdat)
library(DataExplorer)
library(gtsummary)
library(janitor)
library(scales)
library(patchwork)
library(ggrepel)

theme_set(theme_minimal(base_size = 13))
```

```{r}
patients      <- read_csv("../data/raw/patient.csv")
encounter     <- read_csv("../data/raw/encounter.csv")
pulseoximeter <- read_csv("../data/raw/pulseoximeter.csv")
bloodgas      <- read_csv("../data/raw/bloodgas.csv")
```

```{r}
# Cleaned with William's support 
oximetry <- pulseoximeter |>
  inner_join(
    bloodgas |> select(encounter_id, sample, so2),
    by = c("encounter_id", "sample_number" = "sample")
  ) |>
  rename(SpO2 = saturation, SaO2 = so2) |>
  left_join(
    encounter |> select(
      encounter_id, patient_id, fitzpatrick, age_at_encounter
    ),
    by = "encounter_id"
  ) |>
  left_join(
    patients |> select(patient_id, race, ethnicity),
    by = "patient_id"
  ) |>
  mutate(
    bias             = SpO2 - SaO2,
    occult_hypoxemia = SpO2 >= 88 & SaO2 < 88,

    skin_group = cut(
      fitzpatrick,
      breaks         = c(0, 2, 4, 6),
      labels         = c("Light (I–II)", "Medium (III–IV)", "Dark (V–VI)"),
      include.lowest = TRUE
    ),

    race_eth = case_when(
      ethnicity == "Hispanic"                            ~ "Hispanic/Latino",
      str_detect(race, "African American") &
        ethnicity == "Not Hispanic"                      ~ "Black",
      race == "Caucasian" &
        ethnicity == "Not Hispanic"                      ~ "White",
      str_detect(race, "^Asian") &
        ethnicity == "Not Hispanic"                      ~ "Asian",
      TRUE                                               ~ NA_character_)) |>
  filter(!is.na(race_eth))

# Remove non-finite bias values caused by bad join rows
oximetry <- oximetry |>
  filter(is.finite(bias), is.finite(SpO2), is.finite(SaO2))
```

```{r}
#Aestethics purposes 
skin_pal <- c(
  "Light (I–II)"    = "#E8C97A",
  "Medium (III–IV)" = "#B06840",
  "Dark (V–VI)"     = "#4A1E0E")

# Base theme applied to every plot
theme_eda <- function() {
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 14, face = "bold", margin = margin(b = 4)), 
        plot.subtitle = element_text(size = 11, color = "gray"),
        plot.caption = element_text(size = 9,  color = "darkgray"),
        axis.title = element_text(size = 11, color = "darkgray"),
        axis.text = element_text(size = 10, color = "darkgray"),
        panel.grid.major.x  = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 11, face = "bold", color = "darkgray"))}
```

## OpenOximetry Dataset

This group project focuses on analyzing the implications of inaccuracies
in standard pulse oximeters. In this regard, through the analysis of
relevant data, we have observed that oximeters, which function to detect
blood oxygen levels, tend to exhibit wider margins of error for
individuals with dark skin compared to those with light skin.
Scientifically, this phenomenon is explained by the fact that dark skin
contains a layer of melanin that impedes the unimpeded passage of the
oximeter's light beams through the skin, unlike in individuals with
light skin, who lack these elevated levels of melanin. However, as a
group, we believe this issue carries broader implications, and we intend
to analyze how this phenomenon manifests across various demographic
groups. In my specific analysis, I will focus on examining the biases
associated with oximeter models and skin tone, specifically, the biases
that arise among patients of different skin tones, and the interplay of
these factors with respect to sex and race.

## Analysis 1: Bias by device model and skin tone

Given that we are examining potential biases related to skin tone, it is
crucial to consider the devices themselves. In this regard, my aim is to
explore the variability in blood oxygen detection across individual
devices, and to analyze whether there is a general pattern,
specifically, if certain devices are less effective and thus skewing the
overall data results across different skin colors, or if this is a
generalized issue affecting all devices. This is a controlled within
encounter comparison: same patient, only the device differs. Any bias
difference between devices on the same sample is purely a device effect,
not a patient effect.

```{r}
oximetry <- oximetry |>
  mutate(
    device_brand = as.integer(floor(device)),
    device_label = paste0("Model ", as.integer(floor(device))))

# Check: how many distinct brands?
oximetry |> count(device_brand, sort = TRUE)
top_brands <- oximetry |>
  count(device_brand) |>
  filter(n >= 500) |>
  pull(device_brand)

oximetry_top <- oximetry |>
  filter(device_brand %in% top_brands) |>
  mutate(device_label = paste0("Model ", device_brand), device_label = fct_reorder(device_label, device_brand))
```

This is the total amount of all the devices that exist in this dataset,
under the colum "device_brand", togther with the number of dispositives
for each one of the models, as seen, the most common model is the "59",
while the least common model is the "21". As now, this does not tells us
anything directly about the data, but is information that will become
relevant while analyzing all the other devices and the the different
skin colors.

```{r}
# Same device disagreen within different samples 
device_spread <- oximetry_top |>
  filter(between(bias, -20, 20), !is.na(skin_group)) |>
  group_by(encounter_id, sample_number, skin_group) |>
  summarise(
    bias_range = max(bias) - min(bias),
    n_devices  = n_distinct(device_brand),
    .groups    = "drop") |>
  filter(n_devices >= 2)

p_spread <- device_spread |>
  ggplot(aes(x = skin_group, y = bias_range, fill = skin_group)) +
  geom_violin(alpha = 0.45, trim = TRUE, linewidth = 0.3) +
  geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.85, linewidth = 0.4) +
  stat_summary(
    fun = median, geom = "text",
    aes(label = sprintf("%.1f pp", after_stat(y))),
    vjust = -0.5, size = 3.2, color = "black") +
  scale_fill_manual(values = skin_pal) +
  scale_y_continuous(
    labels = function(x) sprintf("%+.0f pp", x),
    limits = c(0, NA)) + 
  labs(
    title    = "Device disagreement at identical sample moments",
    subtitle = "Range of bias readings across devices on the same blood draw",
    x = NULL, y = "Bias range across devices (pp)",
    caption  = "Only samples where ≥ 2 devices present") +
  theme_eda() 

p_spread
```

This visualization highlights a troubling lack of consistency in medical
device readings when used on patients with darker skin tones. As we can
see, there is a range of disagreement between two or more devices
measuring the exact same blood sample at the same moment. While all
groups show some level of discrepancy, the "Dark" skin tone category
(V–VI) has a noticeably higher median disagreement of 4.1 pp compared to
just 3.0 pp for light skin and 2.1 pp for medium skin. More importantly,
the "tail" of the dark skin plot stretches significantly higher,
reaching extremes of over +30 pp in bias range. This means that for
patients with the darkest skin, two different sensors are far more
likely to give wildly conflicting results, creating a dangerous level of
clinical uncertainty that isn't as prevalent for lighter-skinned
patients.

```{r}
# Looking at each individual oxymeter
device_skin_summary <- oximetry_top |>
  filter(!is.na(skin_group), between(bias, -20, 20)) |>
  group_by(device_label, skin_group) |>
  summarise(mean_bias = mean(bias, na.rm = TRUE), n = n(), .groups = "drop")

p_heatmap <- device_skin_summary |>
  ggplot(aes(x = skin_group, y = device_label, fill = mean_bias)) +
  geom_tile(color = "white", linewidth = 1.6) +
  geom_text(aes(label = sprintf("%+.2f", mean_bias), color = abs(mean_bias) > 2), size = 3.6, fontface = "bold", show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "white", "FALSE" = "black")) +
  scale_fill_gradient2(
    low      = "blue",
    mid      = "white",
    high     = "red",
    midpoint = 0,
    limits   = c(-5, 5),
    name     = "Mean bias\n(pp)",
    guide    = guide_colorbar( barwidth = 0.9, barheight = 10, title.position = "top")) +
  labs(title    = "Pulse oximeter bias by device model and skin tone",
       subtitle = "Mean (SpO2 − SaO2) in percentage points · Red = overestimates true oxygen level",
       x = "Fitzpatrick skin tone group", y = "Device model") +
theme_eda() 

p_heatmap
```

This heatmap illustrates how pulse oximeter accuracy varies across
different device models and skin tones, specifically highlighting the
"mean bias" or the difference between the device's reading (Sp02) and
actual blood oxygen levels (Sa02) with the color scale indicating
whether a device tends to overestimate (red/orange) or underestimate
(blue/purple) the true oxygen saturation. A quick look at the darker
skin tone column reveals a noticeable trend toward more intense orange
and red hues, suggesting that many of these models consistently
overestimate oxygen levels in individuals with darker skin. This
discrepancy is critical because an overestimation could potentially mask
dangerously low oxygen levels, highlighting a significant technical bias
in how these medical devices perform across diverse populations.
Therefore, because the majority of them present this orange/red color,
it could be said that this seems to be a problem that the big majority
of the devices have in common and it is not just one device affecting
the entire dataset, which reafirms the importance of this problem since
it is common across all devices.

## Analysis 2: Within patient bias stability across samples

On that note, analyzing within-patient bias stability is also essential
because it helps us distinguish between a device that is consistently
off and one that is dangerously unpredictable. If a pulse oximeter has a
high mean bias but stays stable, a clinician can at least anticipate the
error; however, high volatility within a single session, measured here
through standard deviation and range, suggests that the device is
erratic. This "jitter" is particularly risky in a clinical setting
because it can lead to intermittent misdiagnosis, where a patient might
appear stable during one check but be in respiratory distress the next.

```{r}
within_patient <- oximetry_top |>
  filter(between(bias, -20, 20)) |>
  group_by(patient_id, race_eth, skin_group, device_label) |>
  summarise(n_readings = n(), mean_bias  = mean(bias, na.rm = TRUE), sd_bias = sd(bias,   na.rm = TRUE), .groups    = "drop") |>
  filter(n_readings >= 3, !is.na(sd_bias))
```

```{r}
p_scatter_risk <- within_patient |>
  filter(!is.na(skin_group)) |>
  ggplot(aes(x = mean_bias, y = sd_bias, color = skin_group)) +
  geom_point(alpha = 0.22, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_vline(xintercept = 0,  linetype = "dashed", color = "darkgray", linewidth = 0.4) +
  geom_hline(yintercept = 0,  linetype = "dashed", color = "darkgray", linewidth = 0.4) +
  scale_color_manual(values = skin_pal) +
  scale_x_continuous(labels = function(x) sprintf("%+.0f pp", x)) +
  labs(
    title    = "Bias magnitude vs. bias variability per patient",
    x = "Mean bias (pp)", y = "SD of bias (pp)",
    color = "Skin tone") +
  theme_eda() 

p_scatter_risk
```

As you can see in this scatter plot, this examines the relationship
between how much a device is off on average (mean bias) and how much
that error fluctuates during a single session (SD of bias), providing a
clear picture of device reliability across different skin tones.
Interestingly, while the trend lines show that variability generally
decreases as mean bias moves into positive territory, the distribution
of data points reveals a troubling pattern for patients in the "Dark"
skin tone group. These points are notably more scattered toward the
right side of the x-axis, indicating that for these individuals, devices
are not just more likely to overestimate oxygen levels, but they often
do so with a wide range of inconsistency. This combination of high
positive bias and high variability is the "worst-case scenario" for
clinical safety, as it suggests that the readings for patients with
darker skin are both consistently over-optimistic and technically
unstable, making it significantly harder for medical staff to trust the
data during critical moments of care. This is another factor imporant to
consider in our analysis since it contributes with the fact of knowing
that patients with darker skin colors are generally worse off than the
patients with lighter skin colors.

## Analysis 3: Sex and Race Interactions Between Patients 

At the same time, I am also interesrted in analyzing the intersection of
sex and race with regards to this topic. For me, this is critical
because it uncovers whether clinical disadvantages are compounded for
specific subgroups, such as Black or Hispanic women, rather than just
affecting racial groups uniformly. Pulse oximetry performance can be
influenced by physiological differences, like skin thickness, peripheral
perfusion, and hemoglobin level, that often vary by both sex and
ethnicity. However, the patriarchal society we currently live on has
also influenciated the way in which women, specially women of color, are
treated in medical settings. As a woman of color myself, I want to
perform this analysis by calculating the mean bias and occult hypoxemia
rates across these intersecting identities, seeing if we can identify if
certain patients face a "double burden" of inaccuracy that simple,
single variable analyses might miss. This level of detail is necessary
to determine if current medical device algorithms are failing certain
populations more than others, which is a vital step in addressing
systemic health inequities and ensuring that life saving interventions
are triggered reliably for every patient regardless of their demographic
profile.

```{r}
if (!"assigned_sex" %in% names(oximetry)) {
  oximetry  <- oximetry |>
    left_join(patients |> select(patient_id, assigned_sex), by = "patient_id")}

sex_race <- oximetry |>
  filter(!is.na(race_eth), !is.na(assigned_sex), assigned_sex %in% c("Female", "Male"), between(bias, -20, 20)) |>
  group_by(race_eth, assigned_sex) |>
  summarise(
    mean_bias = mean(bias, na.rm = TRUE),
    se_bias = sd(bias, na.rm = TRUE) / sqrt(n()),
    oh_rate = mean(occult_hypoxemia, na.rm = TRUE),
    n = n(), .groups   = "drop") |>
  filter(n >= 20)

sex_pal <- c("Female" = "magenta", "Male" = "blue")
```

```{r}
p_sex_bias <- sex_race |>
  mutate(race_eth = fct_reorder(race_eth, mean_bias, max)) |>
  ggplot(aes(x = race_eth, y = mean_bias, fill = assigned_sex)) +
  geom_col(position = position_dodge(width = 0.70), width = 0.60, alpha = 0.90) +
  geom_errorbar(aes(ymin = mean_bias - 2 * se_bias, ymax = mean_bias + 2 * se_bias), position = position_dodge(width = 0.70), width = 0.20, linewidth = 0.5, color = "darkgray") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray", linewidth = 0.4) +
  scale_fill_manual(values = sex_pal, name = "Assigned sex") +
  scale_y_continuous(labels = function(x) sprintf("%+.2f pp", x)) +
  labs(title  = "Mean bias by race and assigned sex",
    subtitle = "Error bars = 95% CI · Does sex compound the racial bias gradient?",
    x = NULL, y = "Mean bias (SpO2 − SaO2, pp)") +
  theme_eda() 

# Occult hypoxemia rate by race and assigned sex

p_sex_oh <- sex_race |>
  mutate(race_eth = fct_reorder(race_eth, oh_rate, max)) |>
  ggplot(aes(x = race_eth, y = oh_rate, fill = assigned_sex)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.88) +
  geom_text(
    aes(label = percent(oh_rate, accuracy = 0.1)),
    position = position_dodge(width = 0.75),
    vjust = -0.5, size = 2.9, color = "darkgray") +
  scale_fill_manual(values = sex_pal, name = "Assigned sex") +
  labs(
    title    = "Occult hypoxemia rate by race and assigned sex",
    x = NULL, y = "Occult hypoxemia rate") +
  theme_eda() 

(p_sex_bias / p_sex_oh) +
  plot_annotation(
    title = "Sex × race interaction in pulse oximeter error",
    theme = theme(plot.title = element_text(size = 15, face = "bold")))
```

As expected, these charts reveal a striking interaction between race and
sex, demonstrating that pulse oximeter inaccuracies are not evenly
distributed across demographics. In the top panel, we see that mean
bias, the degree to which the device overestimates oxygen levels, is
consistently higher for Black and Asian patients compared to White and
Hispanic patients, but the sex-based disparity is most pronounced among
Black individuals, where females experience significantly higher
overestimation than males. This technical error translates directly into
clinical risk, as shown in the bottom panel, where the rate of occult
hypoxemia peaks sharply for Black females at 12.6%. This means that for
more than one in ten Black women, the device reports a "safe" oxygen
level when they are actually experiencing dangerously low saturation.
Collectively, the data suggests that while racial bias is a primary
driver of device failure, being female often compounds this risk,
particularly for Black and Asian patients, potentially leading to
systemic delays in care for these specific groups.
