---
title: "Individual EDA: Hidden Hypoxemia and Its Economic Consequences"
subtitle: "OpenOximetry + NY SPARCS Inpatient Discharge Data"
author: "William Acosta Lora"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
editor: 
  markdown: 
    wrap: 72
---

## Overview

This EDA covers both project datasets and the bridge connecting them:

-   **OpenOximetry** (UCSF Hypoxia Lab) — paired pulse oximeter (SpO₂)
    and arterial blood gas (SaO₂) measurements with objective skin tone
    scores. Used to measure who gets misdiagnosed and by how much.
-   **NY SPARCS Inpatient Discharge 2021** — 103K respiratory discharge
    records from New York State hospitals with charges, costs, length of
    stay, insurance, and severity. Used to measure who pays more and
    arrives sicker.
-   **Bridge analysis** — because the two datasets do not share
    patients, the connection is made at the group level: race-stratified
    summaries from each dataset are joined on race to trace the
    downstream economic consequences of pulse oximeter bias.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(visdat)
library(DataExplorer)
library(gtsummary)
library(janitor)
library(scales)
library(patchwork)
library(ggrepel)

theme_set(theme_minimal(base_size = 13))
```

The analysis is structured as a three-part argument. Part 1 establishes
the clinical reality: pulse oximeters systematically fail darker-skinned
patients, and that failure is measurable and graded by skin tone. Part 2
documents the economic reality: Black and Hispanic/Latino patients pay
more for respiratory care at every age, under every insurance type, at
every severity level. Part 3 asks whether these two realities are
connected and finds that the racial ordering in misdiagnosis rates is
consistent with the racial ordering in costs, and that the charge gap
persists even after the two most obvious confounders (insurance and
severity) are removed.

## Part 1 — OpenOximetry Dataset

### Load and Merge

The OpenOximetry dataset is stored as five relational CSVs. All five are
joined into a single analysis-ready tibble. `age_at_encounter` is pulled
from the encounter table and `race_eth` is constructed with ethnicity
taking priority over race to correctly classify Hispanic patients.

```{r}
#| label: load-oximetry
#| message: false
#| warning: false

patients      <- read_csv("../data/patient.csv")
encounter     <- read_csv("../data/encounter.csv")
pulseoximeter <- read_csv("../data/pulseoximeter.csv")
bloodgas      <- read_csv("../data/bloodgas.csv")

oximetry <- pulseoximeter |>
  inner_join(
    bloodgas |> select(encounter_id, sample, so2),
    by = c("encounter_id", "sample_number" = "sample")
  ) |>
  rename(SpO2 = saturation, SaO2 = so2) |>
  left_join(
    encounter |> select(
      encounter_id, patient_id, fitzpatrick, age_at_encounter
    ),
    by = "encounter_id"
  ) |>
  left_join(
    patients |> select(patient_id, race, ethnicity),
    by = "patient_id"
  ) |>
  mutate(
    bias             = SpO2 - SaO2,
    occult_hypoxemia = SpO2 >= 88 & SaO2 < 88,

    skin_group = cut(
      fitzpatrick,
      breaks         = c(0, 2, 4, 6),
      labels         = c("Light (I–II)", "Medium (III–IV)", "Dark (V–VI)"),
      include.lowest = TRUE
    ),

    race_eth = case_when(
      ethnicity == "Hispanic"                            ~ "Hispanic/Latino",
      str_detect(race, "African American") &
        ethnicity == "Not Hispanic"                      ~ "Black",
      race == "Caucasian" &
        ethnicity == "Not Hispanic"                      ~ "White",
      str_detect(race, "^Asian") &
        ethnicity == "Not Hispanic"                      ~ "Asian",
      TRUE                                               ~ NA_character_
    )
  ) |>
  filter(!is.na(race_eth))

# Remove non-finite bias values caused by bad join rows
oximetry <- oximetry |>
  filter(is.finite(bias), is.finite(SpO2), is.finite(SaO2))

glimpse(oximetry)
```

The merged dataset contains **`r nrow(oximetry) |> comma()` paired
measurements** across **`r n_distinct(oximetry$patient_id)` unique
patients**.

The dataset captures paired measurements taken simultaneously by
multiple pulse oximeter devices and an arterial blood gas draw. Each row
is one device reading matched to one blood gas measurement, so a single
patient encounter generates multiple rows — one per device per sample.
The key outcome variables are `bias` (how much the device overestimates
true saturation) and `occult_hypoxemia` (whether the device reads normal
while the patient is actually hypoxic

### Structure and Missingness

```{r}
#| label: oximetry-visdat
#| fig-alt: "Missingness bar chart for key oximetry variables"

oximetry |>
  select(SpO2, SaO2, bias, fitzpatrick, age_at_encounter,
         race_eth, skin_group, occult_hypoxemia) |>
  plot_missing(
    title   = "Missingness in key OpenOximetry variables",
    ggtheme = theme_minimal(base_size = 12)
  )
```

```{r}
#| label: oximetry-missing
#| message: false

plot_missing(
  oximetry,
  title   = "Missing values per variable — OpenOximetry",
  ggtheme = theme_minimal(base_size = 12)
)
```

------------------------------------------------------------------------

### Patient Demographics

```{r}
#| label: oximetry-demographics
#| message: false

oximetry |>
  distinct(patient_id, .keep_all = TRUE) |>
  select(race_eth, age_at_encounter, fitzpatrick, skin_group) |>
  tbl_summary(
    label = list(
      race_eth           ~ "Race / Ethnicity",
      age_at_encounter   ~ "Age at encounter (years)",
      fitzpatrick        ~ "Fitzpatrick score (1–6)",
      skin_group         ~ "Skin tone group"
    ),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany"
  ) |>
  bold_labels() |>
  modify_caption("**Table 1. Patient demographics — OpenOximetry (unique patients)**")
```

```{r fig.width=10}
#| label: oximetry-race-skin
#| fig-alt: "Bar charts of patients by race and by Fitzpatrick skin tone group"

p_race <- oximetry |>
  distinct(patient_id, .keep_all = TRUE) |>
  count(race_eth) |>
  mutate(race_eth = fct_reorder(race_eth, n)) |>
  ggplot(aes(x = n, y = race_eth, fill = race_eth)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Patients by race / ethnicity", x = "Count", y = NULL)

p_skin <- oximetry |>
  distinct(patient_id, .keep_all = TRUE) |>
  filter(!is.na(skin_group)) |>
  count(skin_group) |>
  ggplot(aes(x = skin_group, y = n, fill = skin_group)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_fill_manual(values = c("#f5c07a", "#c8855a", "#6b3a2a")) +
  labs(title = "Patients by Fitzpatrick group", x = NULL, y = "Count")

p_race + p_skin
```

The sample skews toward medium Fitzpatrick scores (III–IV), with fewer
patients at the extremes of the scale. Estimates for Fitzpatrick I–II
and V–VI are based on smaller subgroups and carry more uncertainty than
estimates for the middle range. This is worth keeping in mind when
interpreting the skin-tone gradients below.

### SpO₂, SaO₂, and Bias Distributions

```{r fig.width=10}
#| label: spo2-sao2
#| fig-alt: "Overlapping density plots of SpO2 and SaO2"

oximetry |>
  select(SpO2, SaO2) |>
  pivot_longer(everything(), names_to = "measure", values_to = "value") |>
  ggplot(aes(x = value, fill = measure)) +
  geom_density(alpha = 0.45) +
  scale_fill_manual(values = c("SaO2" = "#2171b5", "SpO2" = "#ef6548")) +
  geom_vline(xintercept = 88, linetype = "dashed", color = "grey40") +
  annotate("text", x = 87.3, y = 0.15, label = "88% threshold",
           hjust = 1, size = 3.5, color = "grey40") +
  labs(
    title = "SpO₂ (pulse ox) vs. SaO₂ (arterial blood gas)",
    x = "Oxygen saturation (%)", y = "Density", fill = "Measurement"
  )
```

```{r}
#| label: bias-dist
#| fig-alt: "Density plot of pulse oximeter bias capped at plus or minus 20pp"

bias_mean <- mean(oximetry$bias, na.rm = TRUE)

oximetry |>
  filter(between(bias, -20, 20)) |>
  ggplot(aes(x = bias)) +
  geom_density(fill = "#9ecae1", alpha = 0.7) +
  geom_vline(xintercept = 0,         linetype = "solid",  color = "grey30") +
  geom_vline(xintercept = bias_mean, linetype = "dashed", color = "#e34a33") +
  annotate("text", x = bias_mean + 0.3, y = Inf, vjust = 1.5,
           label = paste0("Mean = ", round(bias_mean, 2), " pp"),
           color = "#e34a33", size = 3.8) +
  labs(
    title    = "Distribution of pulse oximeter bias (SpO₂ − SaO₂)",
    subtitle = "Positive = device overestimates true oxygen saturation; capped at ±20 pp for readability",
    x = "Bias (pp)", y = "Density"
  )
```

The mean bias of `r round(bias_mean, 2)` percentage points means the
pulse oximeter reports oxygen saturation about one point higher than the
patient actually has on average. That sounds small, but clinical
decisions — whether to administer supplemental oxygen, whether a patient
is stable for discharge — are often made on differences of one or two
points. A device that consistently flatters the reading shifts those
decision thresholds in ways that disadvantage every patient, but
disadvantage darker-skinned patients most.

```{r}
#| label: bias-skin
#| fig-alt: "Violin and boxplot of bias by Fitzpatrick skin tone group"

oximetry |>
  filter(!is.na(skin_group), between(bias, -20, 20)) |>
  ggplot(aes(x = skin_group, y = bias, fill = skin_group)) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("#f5c07a", "#c8855a", "#6b3a2a")) +
  labs(
    title    = "Pulse oximeter bias by Fitzpatrick skin tone group",
    subtitle = "Capped at ±20 pp; shows bulk of distribution",
    x = "Skin tone group", y = "Bias (SpO₂ − SaO₂, pp)"
  ) +
  theme(legend.position = "none")
```

```{r}
#| label: bias-fitzpatrick-scatter
#| message: false
#| fig-alt: "Scatter of bias vs continuous Fitzpatrick score with loess smoother"

oximetry |>
  filter(!is.na(fitzpatrick), between(bias, -20, 20)) |>
  ggplot(aes(x = fitzpatrick, y = bias)) +
  geom_jitter(alpha = 0.15, width = 0.15, color = "#6b3a2a") +
  geom_smooth(method = "loess", se = TRUE, color = "#e34a33", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = 1:6) +
  labs(
    title    = "Pulse oximeter bias vs. continuous Fitzpatrick score",
    subtitle = "Bias increases continuously with darker skin tone; capped at ±20 pp",
    x = "Fitzpatrick score (1 = lightest, 6 = darkest)", y = "Bias (pp)"
  )
```

Bias magnitude matters because it determines whether a patient crosses
the clinical threshold for treatment. The most consequential form of
error is **occult hypoxemia**: the device reads SpO₂ ≥ 88% (appearing
normal) while true SaO₂ is below 88% (the patient is genuinely hypoxic).
This is not a measurement nuisance — it is the specific failure mode
that causes clinicians to withhold oxygen therapy and delay intervention
from patients who need it.

```{r}
#| label: occult-skin
#| fig-alt: "Bar chart of occult hypoxemia rate by skin tone group"

oximetry |>
  filter(!is.na(skin_group)) |>
  group_by(skin_group) |>
  summarise(
    n        = n(),
    n_occult = sum(occult_hypoxemia, na.rm = TRUE),
    rate     = n_occult / n
  ) |>
  ggplot(aes(x = skin_group, y = rate, fill = skin_group)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)), vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format(),
                     expand  = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("#f5c07a", "#c8855a", "#6b3a2a")) +
  labs(
    title = "Occult hypoxemia rate by Fitzpatrick skin tone group",
    x = "Skin tone group", y = "Rate of occult hypoxemia"
  ) +
  theme(legend.position = "none")
```

The pattern by skin tone group maps directly onto race when the same
data is stratified by self-reported race and ethnicity. Because race and
skin tone are correlated but not identical — not all Black patients have
dark Fitzpatrick scores, and not all patients with dark Fitzpatrick
scores are Black — the race analysis captures a slightly different slice
of the disparity. Together, the two views strengthen the case: the
device fails systematically along both a biological dimension (melanin
concentration) and a social one (race as experienced in a clinical
encounter).

### Occult Hypoxemia by Race

```{r}
#| label: occult-race
#| fig-alt: "Bar chart of occult hypoxemia rate by race"

oximetry |>
  filter(!is.na(race_eth)) |>
  group_by(race_eth) |>
  summarise(rate = mean(occult_hypoxemia, na.rm = TRUE), n = n()) |>
  mutate(race_eth = fct_reorder(race_eth, rate)) |>
  ggplot(aes(x = rate, y = race_eth, fill = race_eth)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)),
            hjust = -0.15, size = 3.8) +
  scale_x_continuous(labels = percent_format(),
                     expand  = expansion(mult = c(0, 0.12))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Occult hypoxemia rate by race / ethnicity",
    x = "Rate of occult hypoxemia", y = NULL
  )
```

Part 1 has established who gets misdiagnosed and by how much. Part 2 now
turns to the economic side of the same question: among patients admitted
to New York State hospitals with respiratory diagnoses, who pays more,
who stays longer, and who arrives sicker — and does that ordering match
what the oximetry data predicts?

## Part 2 — NY SPARCS Inpatient Discharge Dataset

### Load and Clean

SPARCS is pulled directly from the NY Health Data API, filtered
server-side to respiratory diagnoses. No file is saved to disk.

```{r}
#| label: load-sparcs
#| message: false
#| warning: false

sparcs_raw <- read_csv(
  "https://health.data.ny.gov/resource/tg3i-cinn.csv?$where=ccsr_diagnosis_code%20like%20%27RSP%25%27&$limit=500000",
  show_col_types = FALSE
)

sparcs <- sparcs_raw |>
  filter(str_starts(ccsr_diagnosis_code, "RSP")) |>
  mutate(
    length_of_stay = as.numeric(length_of_stay),
    total_charges  = as.numeric(total_charges),
    total_costs    = as.numeric(total_costs),

    # Ethnicity-first: Hispanic patients appear under multiple race codes
    race_eth = case_when(
      ethnicity == "Spanish/Hispanic"                    ~ "Hispanic/Latino",
      race == "Black/African American" &
        ethnicity == "Not Span/Hispanic"                 ~ "Black",
      race == "White" &
        ethnicity == "Not Span/Hispanic"                 ~ "White",
      TRUE                                               ~ NA_character_
    ),

    age_group = case_when(
      age_group %in% c("18 to 29", "30 to 49") ~ "18–49",
      age_group == "50 to 69"                  ~ "50–69",
      age_group == "70 or Older"               ~ "70+",
      TRUE                                     ~ NA_character_
    ),

    insurance = case_when(
      str_detect(payment_typology_1, regex("medicaid", ignore_case = TRUE)) ~ "Medicaid",
      str_detect(payment_typology_1, regex("medicare", ignore_case = TRUE)) ~ "Medicare",
      str_detect(payment_typology_1, regex("private",  ignore_case = TRUE)) ~ "Private",
      str_detect(payment_typology_1, regex("self",     ignore_case = TRUE)) ~ "Self-Pay",
      TRUE                                                                   ~ "Other"
    ),

    severity = factor(
      apr_severity_of_illness,
      levels = c("Minor", "Moderate", "Major", "Extreme")
    ),

    severity_code = apr_severity_of_illness_code,

    died = str_detect(
      patient_disposition,
      regex("expired|died", ignore_case = TRUE)
    )
  ) |>
  filter(!is.na(race_eth))

glimpse(sparcs)
```

After filtering, the dataset contains **`r nrow(sparcs) |> comma()`
discharge records**.

Unlike OpenOximetry, which is a controlled lab dataset with careful
measurement protocols, SPARCS is administrative data — collected at the
point of hospital discharge primarily for billing and regulatory
purposes. The economic variables (charges, costs, length of stay) are
measured precisely, but clinical variables like severity are assigned by
coders working from discharge records rather than by clinicians at the
bedside. That distinction matters for interpreting the bridge: if pulse
oximeter bias causes delayed treatment, the true severity of a patient's
condition at the time of intervention may be worse than the coded APR
severity score reflects — the score captures severity as documented, not
severity as it actually developed.

------------------------------------------------------------------------

### Structure and Missingness

```{r}
#| label: sparcs-visdat
#| fig-alt: "Missingness bar chart for key SPARCS variables"

sparcs |>
  select(race_eth, age_group, gender, insurance, severity,
         length_of_stay, total_charges, total_costs, died) |>
  plot_missing(
    title   = "Missingness in key SPARCS variables",
    ggtheme = theme_minimal(base_size = 12)
  )
```

```{r fig.height=10, fig.width=10}
#| label: sparcs-missing
#| message: false

plot_missing(
  sparcs,
  title   = "Missing values per variable — SPARCS",
  ggtheme = theme_minimal(base_size = 12)
)
```

------------------------------------------------------------------------

### Patient Characteristics

```{r }
#| label: sparcs-demographics
#| message: false

sparcs |>
  select(gender, race_eth, age_group, insurance, severity,
         length_of_stay, total_charges, total_costs) |>
  tbl_summary(
    label = list(
      gender         ~ "Gender",
      race_eth       ~ "Race / Ethnicity",
      age_group      ~ "Age group",
      insurance      ~ "Insurance type",
      severity       ~ "Illness severity",
      length_of_stay ~ "Length of stay (days)",
      total_charges  ~ "Total charges ($)",
      total_costs    ~ "Total costs ($)"
    ),
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany"
  ) |>
  bold_labels() |>
  modify_caption("**Table 2. Patient characteristics — SPARCS respiratory discharges**")
```

### Race, Insurance, and Diagnosis Breakdown

```{r fig.height=10, fig.width=10}
#| label: sparcs-race-insurance
#| fig-alt: "Bar charts showing race and insurance breakdown"

p_race_s <- sparcs |>
  count(race_eth) |>
  mutate(race_eth = fct_reorder(race_eth, n)) |>
  ggplot(aes(x = n, y = race_eth, fill = race_eth)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_x_continuous(labels = comma_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Discharges by race / ethnicity", x = "Count", y = NULL)

p_ins <- sparcs |>
  count(insurance) |>
  mutate(insurance = fct_reorder(insurance, n)) |>
  ggplot(aes(x = n, y = insurance, fill = insurance)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_x_continuous(labels = comma_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Discharges by insurance type", x = "Count", y = NULL)

p_race_s + p_ins
```

```{r}
#| label: sparcs-diagnoses
#| fig-alt: "Horizontal bar chart of top 15 respiratory diagnoses"

sparcs |>
  count(ccsr_diagnosis_description) |>
  slice_max(n, n = 15) |>
  mutate(ccsr_diagnosis_description = fct_reorder(ccsr_diagnosis_description, n)) |>
  ggplot(aes(x = n, y = ccsr_diagnosis_description)) +
  geom_col(fill = "#6baed6", alpha = 0.85) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top 15 respiratory diagnoses", x = "Number of discharges", y = NULL)
```

### Distributions of Economic Outcomes

```{r}
#| label: sparcs-econ-dist
#| message: false
#| fig-height: 7

sparcs |>
  select(length_of_stay, total_charges, total_costs, severity_code) |>
  plot_histogram(
    title   = "Distributions of key economic and severity variables",
    ggtheme = theme_minimal(base_size = 12)
  )
```

### Charges and LOS by Race

```{r}
#| label: charges-race
#| fig-alt: "Boxplots of charges and LOS by race"

p_charges_r <- sparcs |>
  filter(!is.na(total_charges), total_charges > 0) |>
  mutate(race_eth = fct_reorder(race_eth, total_charges, median)) |>
  ggplot(aes(x = race_eth, y = total_charges, fill = race_eth)) +
  geom_boxplot(outlier.alpha = 0.04, alpha = 0.8) +
  scale_y_log10(labels = dollar_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Total charges by race (log scale)",
       x = NULL, y = "Total charges") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))

p_los_r <- sparcs |>
  filter(!is.na(length_of_stay)) |>
  mutate(race_eth = fct_reorder(race_eth, length_of_stay, median)) |>
  ggplot(aes(x = race_eth, y = length_of_stay, fill = race_eth)) +
  geom_boxplot(outlier.alpha = 0.04, alpha = 0.8) +
  scale_y_continuous(limits = c(0, 30)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Length of stay by race (capped at 30d)",
       x = NULL, y = "Days") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))

p_charges_r + p_los_r
```

### Insurance Mix by Race

```{r}
#| label: insurance-race-heatmap
#| message: false
#| fig-alt: "Heatmap of insurance type proportion within each racial group"

sparcs |>
  filter(!is.na(insurance)) |>
  count(race_eth, insurance) |>
  group_by(race_eth) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = insurance, y = race_eth, fill = pct)) +
  geom_tile(color = "white") +
  geom_text(aes(label = percent(pct, accuracy = 1)), size = 3.2) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       labels  = percent_format()) +
  labs(
    title = "Insurance type distribution within each racial group",
    x = "Insurance type", y = NULL, fill = "% within group"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

Two patterns in the SPARCS data are worth carrying forward into the
bridge analysis. First, Black and Hispanic/Latino patients face higher
charges and longer stays than White patients in the raw comparisons —
but age is a major confounder because White respiratory patients skew
older and are predominantly on Medicare. Second, the insurance
distribution differs sharply by race, with Black and Hispanic/Latino
patients far more likely to be on Medicaid or Self-Pay. Both of these
alternative explanations need to be examined and ruled out before the
charge disparity can be attributed to a clinical mechanism like pulse
oximeter bias.

## Part 3 — Bridge Analysis: Hypoxemia Leads to Economic Burden

The two datasets do not share patients. The connection is made at the
**group level**: race-stratified summaries from each dataset are joined
on race to test whether the racial gradient in occult hypoxemia is
consistent with the racial gradient in costs and severity.

> **Hypothesized pathway:** darker skin tone = higher occult hypoxemia
> rate = delayed treatment = higher severity at admission = longer stay
> = higher charges = compounded by insurance type

### Race-Level Bridge Table

```{r}
#| label: build-bridge

oximetry_summary <- oximetry |>
  group_by(race_eth) |>
  summarise(
    n_measurements        = n(),
    occult_hypoxemia_rate = mean(occult_hypoxemia, na.rm = TRUE),
    mean_bias             = mean(bias, na.rm = TRUE),
    .groups = "drop"
  )

sparcs_summary <- sparcs |>
  group_by(race_eth) |>
  summarise(
    n_discharges    = n(),
    median_charges  = median(total_charges,  na.rm = TRUE),
    median_costs    = median(total_costs,    na.rm = TRUE),
    median_los      = median(length_of_stay, na.rm = TRUE),
    mean_severity   = mean(severity_code,    na.rm = TRUE),
    mortality_rate  = mean(died,             na.rm = TRUE),
    pct_unprotected = mean(
      insurance %in% c("Medicaid", "Self-Pay"), na.rm = TRUE
    ),
    .groups = "drop"
  )

bridge <- oximetry_summary |>
  inner_join(sparcs_summary, by = "race_eth")




bridge |>
  mutate(
    occult_hypoxemia_rate = percent(occult_hypoxemia_rate, accuracy = 0.1),
    mean_bias             = round(mean_bias, 2),
    median_charges        = dollar(median_charges),
    median_los            = round(median_los, 1),
    mean_severity         = round(mean_severity, 2),
    mortality_rate        = percent(mortality_rate, accuracy = 0.1),
    pct_unprotected       = percent(pct_unprotected, accuracy = 0.1)
  ) |>
  select(race_eth, occult_hypoxemia_rate, mean_bias,
         mean_severity, median_los, median_charges,
         mortality_rate, pct_unprotected) |>
  knitr::kable(
    col.names = c("Race / Ethnicity", "Occult hypoxemia rate",
                  "Mean bias (pp)", "Mean severity", "Median LOS",
                  "Median charges", "Mortality rate", "% Medicaid/Self-Pay"),
    caption = "**Table 3. Race-level bridge: clinical disparity → economic burden**"
  )

# The clearest statement of the connection: the same racial ordering
# that appears in hypoxemia rates also appears in hospital charges
race_order <- bridge |>
  arrange(occult_hypoxemia_rate) |>
  pull(race_eth)
```

This is an **ecological association** across two independent
populations. The table tests whether racial gradients in hypoxemia and
costs are directionally consistent — not whether one causes the other at
the patient level.

The table above shows the race-level bridge at a glance. Reading across
a row gives the full picture for each group: how often they are
misdiagnosed, how biased the device reading is, how sick they are at
admission, how long they stay, what they are billed, and how financially
exposed they are. Reading down a column shows the racial gradient on
each dimension. The question the rest of Part 3 asks is whether those
gradients move together — and whether they survive the most obvious
alternative explanations.

```{r fig.height=10, fig.width=10}
#| label: insurance-stacked-race
#| fig-alt: "Stacked bar showing insurance mix by race ordered by occult hypoxemia rate"

sparcs |>
  filter(!is.na(insurance)) |>
  count(race_eth, insurance) |>
  group_by(race_eth) |>
  mutate(
    pct      = n / sum(n),
    race_eth = factor(race_eth, levels = race_order)
  ) |>
  ungroup() |>
  ggplot(aes(x = race_eth, y = pct, fill = insurance)) +
  geom_col(position = "fill", alpha = 0.85) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Insurance type by race — ordered by occult hypoxemia risk (low -> high)",
    subtitle = "Groups most at risk of hidden hypoxemia have the least financial protection",
    x = NULL, y = "Proportion of discharges", fill = "Insurance"
  )
```

The insurance breakdown, ordered from lowest to highest occult hypoxemia
risk left to right, illustrates the compounding structure of the
problem. Moving from White to Hispanic/Latino to Black, Medicaid share
rises and Medicare share falls — meaning the groups most likely to be
misdiagnosed are also the groups with the least financial cushion when
that misdiagnosis leads to a more expensive hospitalization. The next
plot asks whether the misdiagnosis gradient and the cost gradient point
in the same direction.

### Hypoxemia Charges Side by Side

```{r fig.width=10}
#| label: hypoxemia-charges-sidebyside
#| fig-alt: "Side by side bar charts showing occult hypoxemia rate and median charges by race"

p_hyp <- bridge |>
  mutate(race_eth = factor(race_eth, levels = race_order)) |>
  ggplot(aes(x = race_eth, y = occult_hypoxemia_rate, fill = race_eth)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = percent(occult_hypoxemia_rate, accuracy = 0.1)),
            vjust = -0.4, size = 4) +
  scale_y_continuous(labels = percent_format(),
                     expand  = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Occult hypoxemia rate by race",
    subtitle = "OpenOximetry — who gets misdiagnosed",
    x = NULL, y = "Occult hypoxemia rate"
  )

p_charges <- bridge |>
  mutate(race_eth = factor(race_eth, levels = race_order)) |>
  ggplot(aes(x = race_eth, y = median_charges, fill = race_eth)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = dollar(median_charges, accuracy = 1)),
            vjust = -0.4, size = 4) +
  scale_y_continuous(labels = dollar_format(),
                     expand  = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Median hospital charges by race",
    subtitle = "SPARCS — who pays more",
    x = NULL, y = "Median total charges"
  )

p_hyp + p_charges +
  plot_annotation(
    title    = "The same racial ordering appears in both misdiagnosis rates and hospital charges",
    subtitle = "Black patients: highest occult hypoxemia rate and highest median charges"
  )
```

The bars above place the two datasets in direct conversation. The
ordering on the left — White lowest, Hispanic/Latino moderate, Black
highest — is the occult hypoxemia gradient from OpenOximetry. The
ordering on the right is the median charge gradient from SPARCS. They
are not identical, which is expected: these are different populations
measured in different settings. But the direction is consistent, and
that directional consistency is the ecological evidence the bridge is
designed to test. The double burden plot below adds the insurance
dimension to the same picture.

### The Double Burden

```{r}
#| label: double-burden
#| fig-alt: "Scatter of occult hypoxemia rate vs pct Medicaid or Self-Pay by race"

bridge |>
  ggplot(aes(x = occult_hypoxemia_rate, y = pct_unprotected,
             size = median_charges, label = race_eth)) +
  geom_point(alpha = 0.85, color = "#6a3d9a") +
  geom_text_repel(size = 4, fontface = "bold") +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_continuous(labels = dollar_format(), range = c(4, 12)) +
  labs(
    title    = "The double burden: hidden hypoxemia risk and financial vulnerability",
    subtitle = "Groups most likely to be misdiagnosed are also least insured",
    x = "Occult hypoxemia rate (OpenOximetry)",
    y = "% on Medicaid or Self-Pay (SPARCS)",
    size = "Median charges"
  )
```

The group-level bridge is suggestive but not conclusive for one
important reason: White respiratory patients in SPARCS are substantially
older than Black and Hispanic/Latino patients, and older patients tend
to arrive sicker and stay longer regardless of any clinical bias. If age
is driving the severity and cost differences, the apparent disparity
could be mostly a demographic artifact. The age-stratified analysis
below addresses this directly by comparing racial groups within the same
age band, removing age as a confound.

### Age-Stratified Analysis

Age confounds the group-level comparison — White respiratory patients
skew older and are predominantly on Medicare. Stratifying by age band
enables a fair within-band comparison. The 103K-row SPARCS extract
provides stable cell counts (3,000–26,000 per race × age band).

```{r}
#| label: sparcs-age-summary

sparcs_age <- sparcs |>
  filter(!is.na(race_eth), !is.na(age_group)) |>
  group_by(race_eth, age_group) |>
  summarise(
    n_discharges   = n(),
    median_charges = median(total_charges,  na.rm = TRUE),
    median_costs   = median(total_costs,    na.rm = TRUE),
    median_los     = median(length_of_stay, na.rm = TRUE),
    mean_severity  = mean(severity_code,    na.rm = TRUE),
    mortality_rate = mean(died,             na.rm = TRUE),
    pct_medicaid   = mean(insurance == "Medicaid", na.rm = TRUE),
    .groups = "drop"
  )

sparcs_age |>
  mutate(
    median_charges = dollar(median_charges),
    median_los     = round(median_los, 1),
    mean_severity  = round(mean_severity, 2),
    mortality_rate = percent(mortality_rate, accuracy = 0.1),
    pct_medicaid   = percent(pct_medicaid,   accuracy = 0.1)
  ) |>
  arrange(age_group, race_eth) |>
  knitr::kable(
    col.names = c("Race", "Age group", "N", "Median charges", "Median costs",
                  "Median LOS", "Mean severity", "Mortality rate", "% Medicaid"),
    caption = "**Table 4. Age-stratified outcomes by race — full SPARCS respiratory extract**"
  )
```

Cell counts range from roughly 3,000 to 26,000 per race × age
combination — large enough that all estimates below are stable. The
18–49 band tests whether disparities exist before age-related
comorbidities accumulate. The 50–69 band is the cleanest
insurance-controlled comparison: old enough for respiratory conditions
to carry real clinical weight, young enough that most patients are not
yet on Medicare. The 70+ band is where both insurance and severity
converge — making any residual gap the hardest to explain away and the
most compelling evidence for a pre-admission clinical mechanism.

```{r}
#| label: charges-age-race
#| fig-alt: "Line plot of median charges by age group and race"

sparcs_age |>
  ggplot(aes(x = age_group, y = median_charges,
             color = race_eth, group = race_eth)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = dollar(median_charges, accuracy = 1)),
                  size = 3.2, show.legend = FALSE) +
  scale_y_continuous(labels = dollar_format()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Median charges by age group and race",
    subtitle = "Black and Hispanic/Latino patients are billed more at every age band",
    x = "Age group", y = "Median total charges", color = "Race / Ethnicity"
  )
```

The charge gap is present at every age band and widens with age. At
18–49 the difference is small — under \$1,500 between any two groups. By
50–69 it has opened to roughly \$8K between Hispanic/Latino and White
patients. By 70+ Black patients are billed \~\$16K more and
Hispanic/Latino patients \~\$13K more than White patients at the same
age. The bar chart below isolates the 50–69 band as the single cleanest
age-controlled comparison, where insurance mix still differs across
races but age no longer does.

```{r fig.width=10}
#| label: charges-severity-controlled
#| fig-alt: "Bar chart showing median charges by race at 50-69, the age-controlled comparison"

# 50–69 is the cleanest comparison: old enough to have respiratory comorbidities,
# young enough that most are not yet on Medicare — insurance mix is most comparable
sparcs_age |>
  filter(age_group == "50–69") |>
  mutate(race_eth = fct_reorder(race_eth, median_charges)) |>
  ggplot(aes(x = race_eth, y = median_charges, fill = race_eth)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = dollar(median_charges, accuracy = 1)),
            vjust = -0.4, size = 4.5) +
  scale_y_continuous(labels = dollar_format(),
                     expand  = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Median charges by race — ages 50 to 69 only",
    subtitle = "Age-controlled comparison: Black and Hispanic/Latino patients pay more at the same life stage",
    x = NULL, y = "Median total charges"
  )
```

Within the 50–69 band alone, Hispanic/Latino patients are billed \~\$8K
more than White patients and Black patients \~\$5K more, after removing
age as a variable. This is the most conservative framing of the
disparity — and it still shows a consistent gap. The Medicaid line plot
below adds the final piece by showing how financial vulnerability tracks
the same racial gradient across all three age bands, and what happens to
it at 70+ when Medicare takes over.

```{r  fig.width=10}
#| label: medicaid-age-race
#| fig-alt: "Line plot of Medicaid share by age group and race"

sparcs_age |>
  ggplot(aes(x = age_group, y = pct_medicaid,
             color = race_eth, group = race_eth)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = percent(pct_medicaid, accuracy = 1)),
                  size = 3.2, show.legend = FALSE) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Medicaid share by age group and race",
    subtitle = "Financial vulnerability is highest at working age for Black and Hispanic/Latino patients",
    x = "Age group", y = "% on Medicaid", color = "Race / Ethnicity"
  )
```

**Key finding:** At 70+, two potential explanations for the charge gap
simultaneously collapse. First, Medicaid coverage converges to near zero
across all groups as patients shift to Medicare (Hispanic/Latino 10%,
Black 4%, White 1%) — so insurance can no longer explain the difference.
Second, severity scores converge tightly at 70+ (White 2.79, Black 2.78,
Hispanic/Latino 2.69) — so illness severity at admission cannot explain
it either. Yet Black patients are billed \~\$16K more than White
patients and Hispanic/Latino patients \~\$13K more, at the same age,
with the same insurance, at the same severity. The residual gap —
unexplained by age, insurance, or recorded severity — is consistent with
a clinical mechanism upstream of hospital admission, such as delayed
treatment from undetected hypoxemia inflating the true disease burden
beyond what severity scores capture at the point of care.

------------------------------------------------------------------------

## Summary of Key Findings

| Finding | Evidence | Dataset |
|------------------------|------------------------|------------------------|
| **Darker skin → higher occult hypoxemia** | Black patients: \~10.5% occult hypoxemia rate vs. \~6% for White; a continuous positive gradient is confirmed by loess smoother across all six Fitzpatrick scores | OpenOximetry |
| **Black and Hispanic/Latino patients pay more at every age** | Charge gap is present at 18–49 (\~\$1K) and widens substantially by 70+ (\~\$16K for Black, \~\$13K for Hispanic/Latino vs. White) | SPARCS |
| **The gap survives controlling for both insurance and severity** | At 70+, Medicaid rates and severity scores nearly equalize across racial groups, yet the charge gap reaches its maximum — ruling out both as primary explanations | SPARCS |
| **Financial vulnerability is greatest at working age** | 70% of Hispanic/Latino and 64% of Black respiratory patients aged 18–49 are on Medicaid vs. 39% of White patients — the highest-risk group is also the least protected | SPARCS |
| **The double burden applies differently by group** | Black patients face both elevated hypoxemia risk (\~10.5%) and Medicaid exposure (\~45%); Hispanic/Latino patients face extreme financial vulnerability (\~55% Medicaid) with moderate hypoxemia risk (\~6.5%), suggesting additional drivers beyond device bias | Bridge |
| **Severity and insurance are insufficient explanations** | The charge gap widens precisely when both converge, pointing to a pre-admission clinical mechanism consistent with undetected hypoxemia | Bridge |

::: callout-limitations
Because the two datasets are not linked at the patient level, all bridge
findings are **ecological associations**, not causal estimates. The
pattern is consistent with the pulse oximeter bias hypothesis but cannot
confirm it.
:::
