---
title: "Individual EDA: Looking at "
subtitle: "NY SPARCS Inpatient Discharge Data"
author: "Nayla Trigueros Ortiz"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
editor: 
  markdown: 
    wrap: 72
---

## Overview

This EDA covers both project datasets and the bridge connecting them: -
**NY SPARCS Inpatient Discharge 2021** — 103K respiratory discharge
records from New York State hospitals with charges, costs, length of
stay, insurance, and severity. Used to measure who pays more and arrives
sicker.

I will analyze clinical severity and outcomes across racial and ethnic
groups using the **NY SPARCS Inpatient Discharge 2021** dataset,
filtered to respiratory diagnoses. The central question is whether
Black, White, and Hispanic/Latino patients arrive at New York State
hospitals with different levels of illness severity, experience
different lengths of stay, and face different in-hospital mortality
rates — and whether those differences persist after controlling for age.

The analysis is structured in three parts. First, I characterize the
overall patient population and examine how severity of illness is
distributed across racial groups. Second, I look at length of stay and
in-hospital mortality as downstream clinical outcomes of that severity.
Third, I stratify all findings by age group to remove the most important
demographic confounder: White respiratory patients skew older, and older
patients tend to arrive sicker regardless of race. And finally, ask
whether the disparities hold within the same life stage.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(DataExplorer)
library(scales)
library(ggrepel)

theme_set(theme_minimal(base_size = 13))
```

## Load/Clean

SPARCS is pulled directly from the NY Health Data API, filtered
server-side to respiratory diagnoses (CCSR codes beginning with "RSP").
No file is saved to disk. Race and ethnicity are combined into a single
`race_eth` variable using an ethnicity-first strategy: Hispanic patients
are classified as Hispanic/Latino regardless of their reported race
code, which is necessary because SPARCS frequently records Hispanic
patients under multiple race categories.

```{r}
#| label: load-sparcs
#| message: false
#| warning: false

sparcs_raw <- read_csv(
  "https://health.data.ny.gov/resource/tg3i-cinn.csv?$where=ccsr_diagnosis_code%20like%20%27RSP%25%27&$limit=500000",
  show_col_types = FALSE
)

sparcs <- sparcs_raw |>
  filter(str_starts(ccsr_diagnosis_code, "RSP")) |>
  mutate(
    length_of_stay = as.numeric(length_of_stay),
    total_charges  = as.numeric(total_charges),
    total_costs    = as.numeric(total_costs),

    race_eth = case_when(
      ethnicity == "Spanish/Hispanic"                          ~ "Hispanic/Latino",
      race == "Black/African American" &
        ethnicity == "Not Span/Hispanic"                       ~ "Black",
      race == "White" &
        ethnicity == "Not Span/Hispanic"                       ~ "White",
      TRUE                                                     ~ NA_character_
    ),

    age_group = case_when(
      age_group %in% c("18 to 29", "30 to 49") ~ "18–49",
      age_group == "50 to 69"                  ~ "50–69",
      age_group == "70 or Older"               ~ "70+",
      TRUE                                     ~ NA_character_
    ),

    insurance = case_when(
      str_detect(payment_typology_1, regex("medicaid", ignore_case = TRUE)) ~ "Medicaid",
      str_detect(payment_typology_1, regex("medicare", ignore_case = TRUE)) ~ "Medicare",
      str_detect(payment_typology_1, regex("private",  ignore_case = TRUE)) ~ "Private",
      str_detect(payment_typology_1, regex("self",     ignore_case = TRUE)) ~ "Self-Pay",
      TRUE                                                                   ~ "Other"
    ),

    severity = factor(
      apr_severity_of_illness,
      levels = c("Minor", "Moderate", "Major", "Extreme")
    ),

    severity_code = apr_severity_of_illness_code,

    risk_of_mortality = factor(
      apr_risk_of_mortality,
      levels = c("Minor", "Moderate", "Major", "Extreme")
    ),

    died = str_detect(
      patient_disposition,
      regex("expired|died", ignore_case = TRUE)
    )
  ) |>
  filter(!is.na(race_eth))

glimpse(sparcs)
```

After filtering to respiratory diagnoses and the three focal racial
groups, the dataset contains **`r nrow(sparcs) |> comma()` discharge
records** from New York State hospitals in 2021. Each row represents one
inpatient discharge. The key clinical variables are `severity` (APR-DRG
severity of illness, a four-level ordinal from Minor to Extreme),
`risk_of_mortality` (the corresponding APR-DRG mortality risk score),
`length_of_stay`, and `died` (whether the patient died during the
admission).

## Structure and Missingness

```{r}
#| label: sparcs-visdat
#| fig-alt: "Missingness bar chart for key SPARCS clinical variables"

sparcs |>
  select(race_eth, age_group, severity, risk_of_mortality,
         length_of_stay, total_charges, died) |>
  plot_missing(
    title   = "Missingness in key SPARCS clinical variables",
    ggtheme = theme_minimal(base_size = 12)
  )
```

Missingness is low across all key variables. `severity` and
`risk_of_mortality` are coded at discharge from the patient record and
are nearly complete. `length_of_stay` and `died` are administrative
fields populated for every inpatient record. The analysis proceeds on
the full dataset without imputation.

## Part 1 — Severity of Illness

### Overall Severity Distribution

```{r}
#| label: severity-dist-stacked
#| fig-alt: "Proportional stacked bar chart of severity by race"

sparcs |>
  filter(!is.na(severity)) |>
  count(race_eth, severity) |>
  group_by(race_eth) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = race_eth, y = pct, fill = severity)) +
  geom_col(position = "fill", alpha = 0.85) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "YlOrRd", direction = 1) +
  labs(
    title    = "Severity of illness distribution by race / ethnicity",
    subtitle = "Darker = more severe; share of Major and Extreme cases differs by group",
    x = NULL, y = "Proportion of discharges", fill = "Severity"
  )
```

```{r}
#| label: severity-mean-bar
#| fig-alt: "Bar chart of mean APR severity score with 95% CI error bars by race"

sparcs |>
  filter(!is.na(severity_code)) |>
  group_by(race_eth) |>
  summarise(
    mean_sev = mean(severity_code, na.rm = TRUE),
    se       = sd(severity_code,  na.rm = TRUE) / sqrt(n()),
    n        = n()
  ) |>
  mutate(
    lo       = mean_sev - 1.96 * se,
    hi       = mean_sev + 1.96 * se,
    race_eth = fct_reorder(race_eth, mean_sev)
  ) |>
  ggplot(aes(x = race_eth, y = mean_sev, fill = race_eth)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2, linewidth = 0.8) +
  geom_text(aes(label = round(mean_sev, 2)), vjust = -0.9, size = 4.2) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Mean APR severity score by race / ethnicity",
    subtitle = "Scale: 1 = Minor, 2 = Moderate, 3 = Major, 4 = Extreme; error bars = 95% CI",
    x = NULL, y = "Mean severity score"
  )
```

The severity distribution and mean score plots tell a consistent story:
Black and Hispanic/Latino patients arrive at the hospital with higher
recorded illness severity than White patients on average. The stacked
bars show this as a shift in the composition of discharges — a larger
share of Major and Extreme cases among Black patients, a larger share of
Minor and Moderate cases among White patients. The mean score bar chart
quantifies the gap, and the confidence intervals — tight given the large
sample size — confirm it is not a sampling artifact.

The key caveat is age. The severity score is assigned by administrative
coders at discharge and reflects the full clinical picture at that
point, but White patients skewing older means they have more years of
accumulated comorbidities. The age-stratified analysis below tests
whether the severity gap is an artifact of that demographic difference
or whether it holds within the same age band.

### Risk of Mortality Distribution

```{r}
#| label: risk-mortality-stacked
#| fig-alt: "Proportional stacked bar chart of APR risk of mortality by race"

sparcs |>
  filter(!is.na(risk_of_mortality)) |>
  count(race_eth, risk_of_mortality) |>
  group_by(race_eth) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = race_eth, y = pct, fill = risk_of_mortality)) +
  geom_col(position = "fill", alpha = 0.85) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "OrRd", direction = 1) +
  labs(
    title    = "APR risk of mortality distribution by race / ethnicity",
    subtitle = "Share of Major and Extreme mortality-risk cases differs across groups",
    x = NULL, y = "Proportion of discharges", fill = "Risk of Mortality"
  )
```

The APR risk of mortality score is a related but distinct variable from
severity — it is calibrated specifically to predict in-hospital death
rather than to summarize overall illness burden. The distribution
mirrors the severity pattern: Black patients show a higher share of
Major and Extreme mortality risk, White patients show a higher share of
Minor risk. Hispanic/Latino patients again fall between the two. Because
risk of mortality is a predictive score derived from diagnosis and
procedure codes, a higher score means the clinical record at discharge
codes as more life-threatening — not simply that the patient arrived
older or sicker due to age.


## Part 2 — Clinical Outcomes

### In-Hospital Mortality

```{r}
#| label: mortality-bar
#| fig-alt: "Bar chart of in-hospital mortality rate by race"

sparcs |>
  filter(!is.na(died)) |>
  group_by(race_eth) |>
  summarise(
    n              = n(),
    n_died         = sum(died, na.rm = TRUE),
    mortality_rate = n_died / n
  ) |>
  mutate(race_eth = fct_reorder(race_eth, mortality_rate)) |>
  ggplot(aes(x = race_eth, y = mortality_rate, fill = race_eth)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = percent(mortality_rate, accuracy = 0.1)),
            vjust = -0.4, size = 4.2) +
  scale_y_continuous(labels = percent_format(),
                     expand  = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "In-hospital mortality rate by race / ethnicity",
    subtitle = "Percentage of respiratory discharges that resulted in death",
    x = NULL, y = "Mortality rate"
  )
```

The raw in-hospital mortality rate follows the same racial ordering as
severity: the group with the highest mean severity score has the highest
mortality rate, and the group with the lowest has the lowest. This
ordering is consistent with the severity findings but does not on its
own establish that race is the driver — age and severity are mediating
variables that must be controlled before any causal claim can be made.
The next plot addresses this directly.

### Length of Stay by Race

```{r}
#| label: los-boxplot
#| fig-alt: "Boxplot of length of stay by race, capped at 30 days"

sparcs |>
  filter(!is.na(length_of_stay)) |>
  mutate(race_eth = fct_reorder(race_eth, length_of_stay, median)) |>
  ggplot(aes(x = race_eth, y = length_of_stay, fill = race_eth)) +
  geom_boxplot(outlier.alpha = 0.04, alpha = 0.8) +
  scale_y_continuous(limits = c(0, 30)) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Length of stay by race / ethnicity (capped at 30 days)",
    subtitle = "Boxes show median and interquartile range; outliers beyond 30 days suppressed",
    x = NULL, y = "Days"
  ) +
  theme(legend.position = "none")
```

### Length of Stay within Severity Level

```{r}
#| label: los-severity-line
#| fig-alt: "Line plot of median length of stay by severity and race"

sparcs |>
  filter(!is.na(length_of_stay), !is.na(severity)) |>
  group_by(race_eth, severity) |>
  summarise(median_los = median(length_of_stay, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = severity, y = median_los, color = race_eth, group = race_eth)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = round(median_los, 1)), size = 3.2, show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Median length of stay by severity level and race / ethnicity",
    subtitle = "Within-severity comparison removes illness burden as a confounder",
    x = "Severity of illness", y = "Median LOS (days)", color = "Race / Ethnicity"
  )
```

The within-severity comparison is the strongest single plot in this
section. By holding severity constant, it removes the most obvious
alternative explanation for the LOS gap — that Black and Hispanic/Latino
patients simply arrive sicker. Within every severity category, Black and
Hispanic/Latino patients spend as many or more days in the hospital as
White patients at the same illness level. The gap is not driven by
severity alone. This raises the question of what else might explain it:
differences in treatment course, delays in response to treatment,
comorbidity burden not fully captured by the APR severity score, or
upstream clinical failures that caused the condition to progress further
before the patient sought care.
### Severity Heatmap by Diagnosis and Race

```{r}
#| label: severity-heatmap
#| fig-height: 6
#| fig-alt: "Heatmap of mean severity score by top diagnosis and race"

top_dx <- sparcs |>
  count(ccsr_diagnosis_description) |>
  slice_max(n, n = 10) |>
  pull(ccsr_diagnosis_description)

sparcs |>
  filter(ccsr_diagnosis_description %in% top_dx, !is.na(severity_code)) |>
  group_by(race_eth, ccsr_diagnosis_description) |>
  summarise(mean_sev = mean(severity_code, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = race_eth, y = ccsr_diagnosis_description, fill = mean_sev)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_sev, 2)), size = 3.2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       limits = c(1, 4)) +
  labs(
    title    = "Mean severity score by race and top 10 diagnoses",
    subtitle = "Comparing same diagnosis across groups controls for case-mix differences",
    x = NULL, y = NULL, fill = "Mean severity"
  ) +
  theme(axis.text.y = element_text(size = 9))
```

The heatmap is a diagnosis-controlled comparison: it asks whether, for
the same respiratory condition, Black and Hispanic/Latino patients
present at higher severity than White patients. Cells in the same row
hold the diagnosis constant, so differences across columns reflect
patient-level variation rather than diagnosis-mix. Where a racial group
shows consistently darker cells across diagnoses — not just for one
condition but for many — the severity gap is more likely to reflect
something about the patient's journey to the hospital (access to care,
delays in treatment, or upstream clinical failures) than the type of
condition they were admitted for.

## Part 3 — Age-Stratified Analysis

Age is the most important confounder in this dataset. White respiratory
patients skew older, and older patients accumulate more comorbidities,
generate higher severity scores, and are more likely to die during
admission. Stratifying by age band tests whether the racial gradients in
severity and outcomes hold within the same life stage — or whether they
disappear once age is controlled.

```{r}
#| label: age-strat-summary

sparcs_age <- sparcs |>
  filter(!is.na(race_eth), !is.na(age_group)) |>
  group_by(race_eth, age_group) |>
  summarise(
    n_discharges   = n(),
    mean_severity  = mean(severity_code,    na.rm = TRUE),
    median_los     = median(length_of_stay, na.rm = TRUE),
    mortality_rate = mean(died,             na.rm = TRUE),
    pct_major_ext  = mean(severity %in% c("Major", "Extreme"), na.rm = TRUE),
    .groups = "drop"
  )

sparcs_age |>
  mutate(
    mean_severity  = round(mean_severity, 2),
    median_los     = round(median_los, 1),
    mortality_rate = percent(mortality_rate, accuracy = 0.1),
    pct_major_ext  = percent(pct_major_ext,  accuracy = 0.1)
  ) |>
  arrange(age_group, race_eth) |>
  knitr::kable(
    col.names = c("Race", "Age group", "N", "Mean severity",
                  "Median LOS", "Mortality rate", "% Major / Extreme"),
    caption = "**Table 2. Age-stratified clinical outcomes by race — SPARCS respiratory discharges**"
  )
```

Cell counts range from several thousand to over twenty thousand per race
× age combination, so all estimates are stable. The 18–49 band tests
whether disparities appear before age-related comorbidities accumulate —
this is the cleanest comparison because older patients have more
accumulated health burden by definition. The 50–69 band captures
patients with real respiratory disease burden but before the Medicare
transition dominates. The 70+ band is where age-related confounding is
strongest — making any residual gap in this group the hardest to explain
away.

### Severity by Age Group and Race

```{r}
#| label: severity-age-line
#| fig-alt: "Line plot of mean severity score by age group and race"

sparcs_age |>
  ggplot(aes(x = age_group, y = mean_severity,
             color = race_eth, group = race_eth)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = round(mean_severity, 2)),
                  size = 3.2, show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Mean APR severity score by age group and race / ethnicity",
    subtitle = "Age-controlled comparison: does the severity gap hold within the same life stage?",
    x = "Age group", y = "Mean severity score (1–4)", color = "Race / Ethnicity"
  )
```

### Mortality Rate by Age Group and Race

```{r}
#| label: mortality-age-line
#| fig-alt: "Line plot of in-hospital mortality rate by age group and race"

sparcs_age |>
  ggplot(aes(x = age_group, y = mortality_rate,
             color = race_eth, group = race_eth)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = percent(mortality_rate, accuracy = 0.1)),
                  size = 3.2, show.legend = FALSE) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "In-hospital mortality rate by age group and race / ethnicity",
    subtitle = "Each band holds age constant; differences across lines reflect racial variation",
    x = "Age group", y = "Mortality rate", color = "Race / Ethnicity"
  )
```

### % Major or Extreme by Age Group and Race

```{r}
#| label: major-ext-age-line
#| fig-alt: "Line plot of share of Major or Extreme severity discharges by age group and race"

sparcs_age |>
  ggplot(aes(x = age_group, y = pct_major_ext,
             color = race_eth, group = race_eth)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = percent(pct_major_ext, accuracy = 1)),
                  size = 3.2, show.legend = FALSE) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Share of Major / Extreme severity discharges by age group and race",
    subtitle = "Higher share = more patients arriving at the most severe illness levels",
    x = "Age group", y = "% Major or Extreme", color = "Race / Ethnicity"
  )
```

The age-stratified plots are the core evidence of this section. Across
all three age bands, the racial ordering on severity, mortality, and
share of Major/Extreme cases is consistent: Black patients at or near
the top, White patients at or near the bottom, Hispanic/Latino patients
in between. The gaps narrow somewhat at 70+, where severity scores
converge as all groups accumulate serious illness burden — but they do
not disappear. At 18–49, the comparison is especially striking: these
are working-age patients with fewer accumulated comorbidities, yet Black
and Hispanic/Latino patients still arrive at higher severity levels and
die at higher rates than White patients of the same age.

The persistence of the severity and mortality gaps after controlling for
age points to something happening before the patient reaches the
hospital. One hypothesis — consistent with the broader group project —
is that clinical tools used earlier in the care pathway are less
accurate for darker-skinned patients, allowing conditions to deteriorate
further before they are detected and treated. That upstream failure,
whatever its source, would produce exactly the pattern seen here: higher
severity at admission, longer stays, and higher mortality — not because
these patients are intrinsically sicker, but because their illness was
permitted to progress further before intervention.

## Summary of Key Findings

| Finding | Evidence |
|------------------------------------|------------------------------------|
| **Black patients arrive at higher severity** | Higher mean APR severity score and larger share of Major/Extreme discharges than White or Hispanic/Latino patients in raw comparisons |
| **The severity gap is not fully explained by diagnosis mix** | Heatmap shows higher mean severity for Black patients within the same respiratory diagnosis |
| **LOS gap persists within severity level** | At every severity category, Black and Hispanic/Latino patients have equal or longer median stays than White patients at the same illness level |
| **Racial ordering in mortality is consistent** | In-hospital mortality rate follows the same ordering as severity: Black ≥ Hispanic/Latino ≥ White |
| **Disparities hold after controlling for age** | Age-stratified analysis shows the severity and mortality gap is present at 18–49, 50–69, and 70+, ruling out age as the primary explanation |
| **The 18–49 band is the sharpest evidence** | Among working-age patients with the fewest accumulated comorbidities, racial gaps in severity and mortality are still present — pointing to a pre-admission mechanism |

::: callout-note
SPARCS is administrative data collected at discharge for billing and
regulatory purposes. Severity scores are assigned by coders working from
discharge records rather than by clinicians at the bedside. If earlier
clinical failures cause a condition to deteriorate undetected before
admission, the true disease burden may be understated by the coded
severity score — meaning the observed disparities may actually
underestimate the gap in illness progression before hospital arrival.
:::
