---
title: "Project Proposal"
subtitle: "Undetected and Undertreated: Racial and Gender Disparities in Hidden Hypoxemia and Their Economic Consequences"
---

## Team Members

- **Nayla Trigueros Ortiz** *(Style Lead)*
- **Patricia Escobar Contreras** *(Point of Contact)*
- **William Acosta Lora** *(Technical Lead)*

---

## Project Description

We are interested in exploring racial and gender disparities in access, treatment, and healthcare outcomes in the medical industry for patients. Specifically, we focus on **hidden hypoxemia** — a condition where pulse oximeters fail to accurately detect low blood oxygen levels in patients with darker skin tones — as a measurable, concrete case study of how medical device bias translates into unequal care and unequal economic burden for patients of color.

Our project combines two datasets: the **OpenOximetry dataset** (UCSF Hypoxia Lab), which gives us paired pulse oximeter and blood gas measurements stratified by skin tone and race, and the **NY SPARCS Inpatient Discharge dataset**, which provides hospital-level economic outcomes (total charges, length of stay, costs, mortality) broken down by race, gender, and diagnosis. Together, they let us connect the clinical disparity to its downstream economic consequences.

---

## Research Questions

1. **How does pulse oximetry accuracy vary across skin tone, race, and gender, and what is the magnitude of hidden hypoxemia disparities?**
   Using the OpenOximetry dataset, we will quantify how much pulse oximeters overestimate blood oxygen saturation across Fitzpatrick and Monk skin tone groups and self-reported race and ethnicity, and whether gender interacts with these measurement disparities.

2. **What are the economic consequences of respiratory misdiagnosis and delayed treatment for patients of color in hospital settings?**
   Using the NY SPARCS dataset filtered to respiratory diagnoses, we will compare total hospital charges, costs, length of stay, and discharge disposition across racial and ethnic groups, controlling for severity of illness and insurance type.

3. **Do race, gender, and insurance status intersect to compound economic disparities in respiratory care?**
   We will explore whether the economic burden of poor respiratory care outcomes is compounded by the intersection of race, gender, and payer type — for example, whether Black women on Medicaid face systematically higher costs and longer stays than white men with private insurance for the same respiratory diagnosis severity.

---

## Motivation and Inspiration

As international students and people of color, we are really interested in the differences between the medical care received for darker and lighter-skinned people. Regardless of coming from different countries across Latin America, we have all been able to discuss and realize some of the differences regarding the mistreatments that exist within a predominantly white country, and we want to explore the implications that this has in the medical sector.

Hidden hypoxemia through pulse oximeter bias is a particularly powerful entry point because it is not a question of patient behavior or lifestyle — it is a direct, measurable failure of medical technology. Research has documented that Black and Hispanic patients are significantly more likely to experience undetected hypoxemia, which has been linked to delayed eligibility for critical treatments and worse outcomes. The economic consequences of this — higher costs, longer stays, greater morbidity — fall disproportionately on communities of color, making this both a health equity and an economic justice issue.

---

## Dataset

We use two complementary datasets that together let us measure both the clinical disparity (who gets misdiagnosed) and the economic consequences (who pays more and stays longer).

### NY SPARCS Hospital Inpatient Discharges (2021)

The **New York State Statewide Planning and Research Cooperative System (SPARCS)** is a free, publicly available hospital discharge database from the NY Department of Health. The 2021 de-identified file contains over 2 million patient-level discharge records with race, ethnicity, gender, diagnosis, total charges, total costs, length of stay, insurance type, and severity of illness.

The dataset is downloaded directly from the NY Health Data open portal — no login or account needed. We load it locally and filter to respiratory diagnoses only using the `ccsr_diagnosis_code` column, which uses CCSR (Clinical Classifications Software Refined) codes. All respiratory diagnoses begin with `"RSP"`.

```{r}
#| label: load-sparcs
#| message: false
#| warning: false

library(tidyverse)

# Source: https://health.data.ny.gov/resource/tg3i-cinn.csv?$limit=10000
# Downloaded from NY Health Data open portal 

sparcs_raw <- read_csv("data/tg3i-cinn.csv", show_col_types = FALSE)

# Filter to respiratory diagnoses only (all CCSR codes starting with RSP)
# and clean numeric columns
sparcs <- sparcs_raw |>
  filter(str_starts(ccsr_diagnosis_code, "RSP")) |>
  mutate(
    length_of_stay = as.numeric(length_of_stay),
    total_charges  = as.numeric(total_charges),
    total_costs    = as.numeric(total_costs)
  )

glimpse(sparcs)
```

```{r}
#| label: sparcs-diagnosis-breakdown
#| message: false
#| warning: false

# Respiratory diagnosis codes present in the data
sparcs |>
  count(ccsr_diagnosis_code, ccsr_diagnosis_description) |>
  arrange(desc(n))
```

```{r}
#| label: sparcs-race-breakdown
#| message: false
#| warning: false

# Racial and ethnic breakdown of respiratory patients
sparcs |>
  count(race, ethnicity) |>
  arrange(desc(n))
```

**Key variables:**

| Variable | Description |
|---|---|
| `race` / `ethnicity` | Patient self-reported race and Hispanic/Latino ethnicity |
| `gender` | Patient sex |
| `length_of_stay` | Days hospitalized |
| `total_charges` | Amount billed by the hospital |
| `total_costs` | Estimated actual cost of care |
| `payment_typology_1` | Insurance type (Medicaid, Medicare, Private Health Insurance, Self-Pay) |
| `apr_severity_of_illness_code` | 1–4 scale of illness severity at admission |
| `apr_severity_of_illness` | Severity label: Minor, Moderate, Major, Extreme |
| `apr_risk_of_mortality` | Risk label: Minor, Moderate, Major, Extreme |
| `ccsr_diagnosis_code` | CCSR diagnosis code — RSP codes = respiratory diagnoses |
| `ccsr_diagnosis_description` | Plain-text diagnosis name |
| `patient_disposition` | Outcome: home, expired, transferred, etc. |
| `hospital_county` | County of the treating hospital |
| `emergency_department_indicator` | Whether patient came through the ER |

**Source:** NY State Department of Health — Health Data NY  
**Download URL:** <https://health.data.ny.gov/resource/tg3i-cinn.csv?$limit=10000>
**License:** Public domain / open government data — no login required

---

### OpenOximetry Dataset (UCSF Hypoxia Lab)

The **OpenOximetry dataset** is a prospective, open-access dataset from the University of California San Francisco Hypoxia Lab. It contains paired SpO₂ (pulse oximeter) and SaO₂ (arterial blood gas gold standard) measurements alongside objective skin tone measurements using the Fitzpatrick scale, Monk scale, and spectrophotometry. This dataset is stored locally as five relational CSVs.

```{r}
#| label: load-openoximetry
#| message: false
#| warning: false
#| eval: false

# Load locally stored OpenOximetry tables (place CSVs in /data folder)
patients          <- read_csv("data/patient.csv")
encounter         <- read_csv("data/encounter.csv")
pulseoximeter     <- read_csv("data/pulseoximeter.csv")
bloodgas          <- read_csv("data/bloodgas.csv")
spectrophotometer <- read_csv("data/spectrophotometer.csv")

# Merge tables and compute key outcome variables
oximetry <- pulseoximeter |>
  inner_join(
    bloodgas |> select(encounter_id, sample, so2, ph, pco2, po2, thb),
    by = c("encounter_id", "sample_number" = "sample")
  ) |>
  rename(SpO2 = saturation, SaO2 = so2) |>
  left_join(
    encounter |> select(
      encounter_id, patient_id, age_at_encounter,
      fitzpatrick, monk_forehead, monk_dorsal,
      monk_palmar, monk_upper_arm
    ),
    by = "encounter_id"
  ) |>
  left_join(
    patients |> select(patient_id, assigned_sex, race, ethnicity),
    by = "patient_id"
  ) |>
  mutate(
    bias             = SpO2 - SaO2,
    occult_hypoxemia = SpO2 >= 88 & SaO2 < 88,
    skin_group       = cut(
      fitzpatrick,
      breaks         = c(0, 2, 4, 6),
      labels         = c("Light", "Medium", "Dark"),
      include.lowest = TRUE
    )
  )

glimpse(oximetry)
```

**Key variables:** `SpO2`, `SaO2`, `bias` (SpO2 − SaO2), `occult_hypoxemia`, `fitzpatrick` scale (1–6), Monk scale scores by body location, `race`, `ethnicity`, `assigned_sex`

**Source:** OpenOximetry Project, UCSF Hypoxia Lab  
**Paper:** <https://www.nature.com/articles/s41597-025-04870-8>
**Access:** Data Use Agreement required at the OpenOximetry portal

---

## Communication Plan

- **Primary channels:** WhatsApp and Google Chat
- **Expected response time:** Within 1 business day on weekdays
- **Standing meeting:** Thursdays, 3:00–4:30 PM (as needed throughout the project)

---

## Roles

| Role | Team Member |
|---|---|
| **Point of Contact** — emails instructor and preceptors | Patricia Escobar Contreras |
| **Technical Lead** — oversees codebase and Quarto website | William Acosta Lora |
| **Style Lead** — enforces COMP/STAT 212 Code Styling Guide | Nayla Trigueros Ortiz |

All three team members are expected to contribute equally to data analysis, visualization, and writing.

---

## Conflict Resolution

Communication will be key. If any of the members will be late to a commitment or misses a deadline, they should feel comfortable communicating that to the team and we will redistribute the workload accordingly. If anyone has to take on more work for a deadline due to a re-adjustment, we will adjust back for the next deadline so that the responsibilities remain balanced.

For disagreements where the team is split, we will discuss openly during our Thursday meeting. Ties on project direction will be resolved by a short async team vote with a 24-hour window, deferring to the lead most relevant to the decision (e.g., technical disagreements go to the Technical Lead, formatting to the Style Lead).

---

## Implementation Plan

The following tasks will also be added as **GitHub Issues** in the project repository, with role and priority labels.

| # | Task | Due | Lead |
|---|---|---|---|
| 1 | Set up GitHub repo and Quarto project website skeleton | Week 1 | William |
| 2 | Submit this proposal | Week 2 | Patricia |
| 3 | EDA on SPARCS: charges, LOS, and mortality by race, gender, and insurance | Week 3 | Patricia + William |
| 4 | EDA on OpenOximetry: SpO₂ bias and occult hypoxemia rates by skin tone and race | Week 3 | William + Nayla |
| 5 | Literature review write-up: clinical and economic evidence for hidden hypoxemia disparities | Week 3–4 | Nayla |
| 6 | Visualizations: cost and LOS disparities by race and gender from SPARCS | Week 4 | Patricia |
| 7 | Visualizations: bias distributions and occult hypoxemia rates by skin tone from OpenOximetry | Week 4 | William |
| 8 | Intersectional analysis: race x gender x insurance interaction effects on costs | Week 5 | All |
| 9 | Draft results and interpretation sections | Week 6 | All |
| 10 | Style pass: code review against COMP/STAT 212 Code Styling Guide | Week 7 | Nayla |
| 11 | Final write-up, website polish, and presentation prep | Week 8 | All |


